---
title: "Github-Data Cleaning"
output: html_document
date: "2025-04-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(haven) #reading .sav (SPSS)
library(dplyr) #data wrangling
library(purrr) #efficient iteration over datasets
```


```{r}
#Load ANES Datasets: 2012, 2016, 2018, 2020 

anes_12 <- read_dta("ANES_04_24_ALL/anes_12.dta")
anes_16 <- read_dta("ANES_04_24_ALL/anes_16.dta")
anes_20 <- read.csv("ANES_04_24_ALL/anes_20.csv")
anes_24 <- read.csv("ANES_04_24_ALL/anes_24.csv")

```

##Creating A Only-Latino Datasets
#anes_12_latino, anes_16_latino, anes_20-latino, anes_24_latino
```{r}
# Filter Latino respondents in each dataset: latino_12, latino_16, latino_20, latino_24

latino_12 <- anes_12 %>%
  filter(dem_hisp == 1)  # 2012 Hispanic identifier

latino_16 <- anes_16 %>%
  filter(V161309 == 1)  # 2016 Hispanic identifier 

latino_20<- anes_20 %>%
  filter(V201549x == 1)  # 2020 Hispanic identifier

latino_24 <- anes_24 %>%
  filter(V241501x == 1)  # 2024 Hispanic identifier
```

2012 ANES dataset: 1005 latino identified respondents
2016 ANES dataset: 450 latino identified respondents ****** look into why the sample is small
2020 ANES dataset: 5963 latino identified respondents
2024 ANES dataset: 2301 latino identified respondents

##Harmonizing Variables 


#Convert V241566x (Household Income) Into a Social Class Variable
ANES 2024 lacks a self-identified social class variable, but has V241566x (Total Household Income)

Since income and self-reported social class are related, you can approximate social class using household income categories.

```{r}
#check distribution of V241566x
table(latino_24$V241566x)
summary(latino_24$V241566x)
```
Remove NA(-9,-5,-4,-1) from Analysis
```{r}
latino_24 <- latino_24 %>%
  filter(latino_24$V241566x > 0)  # Keep only valid responses
```
Understanding the Income Ranges
```{r}
table(latino_24$V241566x)
summary(latino_24$V241566x)
```
there are 28 levels of income. We need to group into ranges. 

-Recode V241566x Into Social Class-

Assuming V241566x represents income brackets, and based on the median (17) and distribution, we approximate the social class categories:

```{r}
latino_24 <- latino_24 %>%
  mutate(class = case_when(
    V241566x %in% 1:7 ~ "Lower class",       # Low-income groups
    V241566x %in% 8:14 ~ "Working class",    # Moderate-income groups
    V241566x %in% 15:21 ~ "Middle class",    # Mid-range income
    V241566x %in% 22:28 ~ "Upper class",     # High-income groups
    TRUE ~ NA_character_  # Assign NA if missing
  ))
```

Verify the Recoding
```{r}
table(latino_24$class)
summary(latino_24$class)
```


#Rename all Variables to Common Format
```{r}
#2012 latino respondent DF
latino_12 <- latino_12 %>%
  rename(
    vote_pres = presvote2012_x,
    partyid = prevote_regpty,
    relig = relig_churchoft,
    poltalk = discuss_disc,
    engskill = dem_hisplang,
    voted = interest_voted2008,
    genstatus = dem_parents,
    age = dem_age_r_x,
    gender = gender_respondent_x,
    educ = dem_edu,
    class = dem_whichclass
  )

#2016 latino respondent DF
latino_16 <- latino_16 %>%
  rename(
    vote_pres = V162034a,
    partyid = V162030,
    relig = V161245,
    poltalk = V161268,
    engskill = V161323,
    voted = V162174,
    genstatus = V161315,
    age = V161267,
    gender = V161342,
    educ = V161270,
    class = V162132
  )

#2020 latino respondent DF
latino_20 <- latino_20 %>%
  rename(
    vote_pres = V202073,
    partyid = V202064,
    relig = V201453,
    poltalk = V202022,
    engskill = V201562,
    voted = V201101,
    genstatus = V201553,
    age = V201507x,
    gender = V201600,
    educ = V201510,
    class = V202352
  )

#2024 latino respondent DF
latino_24 <- latino_24 %>%
  rename(
    vote_pres = V241039,
    partyid = V241227x,
    relig = V241440,
    poltalk = V241004,
    engskill = V241516,
    voted = V241110,
    genstatus = V241506,
    age = V241458x,
    gender = V241551,
    educ = V241463,
    class = class
  )
```

#Convert All Variables to the Correct Type

Modify each dataset (latino_12, latino_16, latino_20, latino_24) before merging:
```{r}
latino_12 <- latino_12 %>%
  mutate(
    vote_pres = as.numeric(vote_pres),
    partyid = as.numeric(partyid),
    relig = as.numeric(relig),
    engskill = as.numeric(engskill),
    poltalk = as.numeric(poltalk),
    voted = as.numeric(voted),
    genstatus = as.numeric(genstatus),
    educ = as.numeric(educ),
    class = as.character(class)  # Convert class to character for uniformity
  )

latino_16 <- latino_16 %>%
  mutate(
    vote_pres = as.numeric(vote_pres),
    partyid = as.numeric(partyid),
    relig = as.numeric(relig),
    engskill = as.numeric(engskill),
    poltalk = as.numeric(poltalk),
    voted = as.numeric(voted),
    genstatus = as.numeric(genstatus),
    educ = as.numeric(educ),
    class = as.character(class)  
  )

latino_20 <- latino_20 %>%
  mutate(
    vote_pres = as.numeric(vote_pres),
    partyid = as.numeric(partyid),
    relig = as.numeric(relig),
    engskill = as.numeric(engskill),
    poltalk = as.numeric(poltalk),
    voted = as.numeric(voted),
    genstatus = as.numeric(genstatus),
    educ = as.numeric(educ),
    class = as.character(class)  
  )

latino_24 <- latino_24 %>%
  mutate(
    vote_pres = as.numeric(vote_pres),
    partyid = as.numeric(partyid),
    relig = as.numeric(relig),
    engskill = as.numeric(engskill),
    poltalk = as.numeric(poltalk),
    voted = as.numeric(voted),
    genstatus = as.numeric(genstatus),
    educ = as.numeric(educ),
    class = as.character(class)  
  )
```

#Keep Only Harmonized Variables

```{r}
common_vars <- c("vote_pres", "partyid", "relig", "poltalk", "engskill", "voted", "genstatus", "age", "gender", "educ", "class")

#2012
latino_12 <- latino_12 %>%
  select(all_of(common_vars)) %>%
  mutate(year = 2012)
#2016
latino_16 <- latino_16 %>%
  select(all_of(common_vars)) %>%
  mutate(year = 2016)
#2020
latino_20 <- latino_20 %>%
  select(all_of(common_vars)) %>%
  mutate(year = 2020)
#2024
latino_24 <- latino_24 %>%
  select(all_of(common_vars)) %>%
  mutate(year = 2024)

```


```{r}
# Combine all datasets into one
latino_all <- bind_rows(latino_12, latino_16, latino_20, latino_24)
```

```{r}
# cleaning out NA
all_clean <- latino_all %>%
  filter(
    !is.na(vote_pres),
    !is.na(partyid),
    !is.na(relig),
    !is.na(poltalk),
    !is.na(engskill),
    !is.na(voted),
    !is.na(genstatus),
    !is.na(age),
    !is.na(gender),
    !is.na(educ),
    !is.na(class)
  )
```

```{r}
write.csv(all_clean, "all_clean.csv", row.names = FALSE)
```

##Make Binary Dependent Variable: Republican Vote 

```{r}
recode_republican_vote <- function(df, year) {
  df %>%
    mutate(republican_vote = case_when(
      year == 2012 & vote_pres == 2 ~ 1,  # Republican (Romney)
      year == 2012 & vote_pres != 2 ~ 0,  # Not Republican
      
      year == 2016 & vote_pres == 2 ~ 1,  # Republican (Trump)
      year == 2016 & vote_pres != 2 ~ 0,  # Not Republican

      year == 2020 & vote_pres == 2 ~ 1,  # Republican (Trump)
      year == 2020 & vote_pres != 2 ~ 0,  # Not Republican

      year == 2024 & vote_pres == 2 ~ 1,  # Republican (TBD)
      year == 2024 & vote_pres != 2 ~ 0,  # Not Republican
      
      vote_pres %in% c(-9, -8, -5, -4, -1) ~ NA_real_,  # Missing values
      TRUE ~ NA_real_
    ))
}
all_clean <- recode_republican_vote(all_clean)
```

```{r}
table(all_clean$year, all_clean$republican_vote, useNA = "ifany")
```


## Create Political Assimilation Based on Indicators 
1. English Proficiency → (Language Spoken at Home)
2. Participation in Political Discussion → (Discussing Politics with Family & Friends)
3. Voter Participation History → (Did You Vote in the Last Election?)
4. Generational Status → (Native Status of Parents)

#Recode All Indicators to Numeric Scales 
#English Proficiency
```{r}
#higher= more english use -> reverse the scale

all_clean <- all_clean %>%
  mutate(
    engskill_recoded = case_when(
      engskill %in% c(1, 2) ~ 5,  # Only/mostly English
      engskill == 3 ~ 3,
      engskill %in% c(4, 5) ~ 1,  # Mostly/only Spanish
      TRUE ~ NA_real_
    )
  )
```

#Political Discussion
```{r}
#need to harmonize across years:
  #2012-2020: 1= Yes, 2 = No -> recode to 1 = frequent, 0= infrequent
  #2024: 1= Always -> reverse to match scale 

all_clean <- all_clean %>%
  mutate(
    poltalk_recoded = case_when(
      year %in% c(2012, 2016, 2020) & poltalk == 1 ~ 1,
      year %in% c(2012, 2016, 2020) & poltalk == 2 ~ 0,
      year == 2024 & poltalk %in% 1:5 ~ 6 - poltalk,  # Reverse so 5=Always
      TRUE ~ NA_real_
    )
  )
```

#Voter Participation (voted)

```{r}
#standardize across the years: 1= Voted, 2= Did not vote

all_clean <- all_clean %>%
  mutate(
    voted_recoded = case_when(
      voted == 1 ~ 1,
      voted == 2 ~ 0,
      TRUE ~ NA_real_
    )
  )
```

#Generational Status (genstatus)

```{r}
#1= U.S born parents -> more assimilated
all_clean <- all_clean %>%
  mutate(
    genstatus_recoded = case_when(
      genstatus == 1 ~ 3,  # Both parents U.S. born
      genstatus == 2 ~ 2,
      genstatus == 3 ~ 1,  # Neither parent U.S. born
      TRUE ~ NA_real_
    )
  )
```

#Standardize Each Indicator
```{r}
# z-score all recoded variables so they contribute equally to the index
all_clean <- all_clean %>%
  mutate(
    z_engskill = scale(engskill_recoded)[,1],
    z_poltalk = scale(poltalk_recoded)[,1],
    z_voted = scale(voted_recoded)[,1],
    z_genstatus = scale(genstatus_recoded)[,1]
  )
```


#Create Political Assimilation Index
```{r}
all_clean <- all_clean %>%
  mutate(
    assimilation_index = rowMeans(select(., z_engskill, z_poltalk, z_voted, z_genstatus), na.rm = TRUE)
  )
```

```{r}
#check index
summary(all_clean$assimilation_index)
hist(all_clean$assimilation_index, main = "Political Assimilation Index", xlab = "Z-score Average")
```

#Standardize Partisan Identification

 Key Adjustments for 2012
Democratic Party (1 in 2012) → Strong/Lean Democrat
Republican Party (2 in 2012) → Strong/Lean Republican
None/Independent (4 in 2012) → Independent
Other (5 in 2012) → Categorized as "Other" or NA
Inapplicable, Don't Know, and Refused (-1, -8, -9) → NA


```{r}

recode_partyid <- function(df, year) {
  df %>%
    mutate(partyid = case_when(
      year == 2012 & partyid == 1 ~ 1,  # Democrat
      year == 2012 & partyid == 2 ~ 2,  # Republican
      year == 2012 & partyid == 4 ~ 3,  # Independent
      year == 2012 & partyid == 5 ~ 4,  # Other
      
      year == 2016 & partyid %in% c(1, 2) ~ 1,  # Democrat
      year == 2016 & partyid %in% c(5, 6) ~ 2,  # Republican
      year == 2016 & partyid %in% c(3, 4) ~ 3,  # Independent
      year == 2016 & partyid == 7 ~ 4,  # Other
      
      year == 2020 & partyid %in% c(1, 2) ~ 1,  # Democrat
      year == 2020 & partyid %in% c(5, 6) ~ 2,  # Republican
      year == 2020 & partyid %in% c(3, 4) ~ 3,  # Independent
      year == 2020 & partyid == 7 ~ 4,  # Other
      
      year == 2024 & partyid %in% c(1, 2) ~ 1,  # Democrat
      year == 2024 & partyid %in% c(5, 6) ~ 2,  # Republican
      year == 2024 & partyid %in% c(3, 4) ~ 3,  # Independent
      year == 2024 & partyid == 7 ~ 4,  # Other
      
      partyid %in% c(-9, -8, -5, -4, -1) ~ NA_real_,  # Set missing values to NA
      TRUE ~ partyid  # Keep original if not recoded
    ))
}

all_clean <- recode_partyid(all_clean)


```

```{r}
table(all_clean$partyid)
```

```{r}
# convert partyid to a factor for logistic regression
all_clean$partyid <- factor(all_clean$partyid,
                                   levels = c(1, 2, 3, 4),
                                   labels = c("Democrat", "Republican", "Independent", "Other"))
```

# Recode Religiosity 

```{r}
recode_religiosity <- function(df) {
  df %>%
    mutate(relig = case_when(
      # Standardized coding for all years
      year %in% c(2012, 2016, 2020, 2024) & relig == 1 ~ 5,  # Every week
      year %in% c(2012, 2016, 2020, 2024) & relig == 2 ~ 4,  # Almost every week
      year %in% c(2012, 2016, 2020, 2024) & relig == 3 ~ 3,  # Once or twice a month
      year %in% c(2012, 2016, 2020, 2024) & relig == 4 ~ 2,  # A few times a year
      year %in% c(2012, 2016, 2020, 2024) & relig == 5 ~ 1,  # Never

      # Handle missing values for all years
      year == 2012 & relig %in% c(-1, -9) ~ NA_real_,
      year == 2016 & relig %in% c(-1, -9, -8, 2) ~ NA_real_,
      year == 2020 & relig %in% c(-9, -8, -1) ~ NA_real_,
      year == 2024 & relig %in% c(-9, -1) ~ NA_real_,

      TRUE ~ relig  # Keep original values if they don't match any case
    ))
}

all_clean <- recode_religiosity(all_clean)
```

```{r}
table(all_clean$relig)
```
#Recode Gender (Binary)

```{r}
recode_gender <- function(df) {
  df %>%
    mutate(gender = case_when(
      # 2012 Scale (1 = Male, 2 = Female)
      year == 2012 & gender == 1 ~ 0,  # Male → 0
      year == 2012 & gender == 2 ~ 1,  # Female → 1
      
      # 2016 Scale (1 = Male, 2 = Female, 3 = Other, -9 = Refused)
      year == 2016 & gender == 1 ~ 0,  # Male → 0
      year == 2016 & gender == 2 ~ 1,  # Female → 1
      year == 2016 & gender == 3 ~ 2,  # Other → 2
      year == 2016 & gender == -9 ~ NA_real_,  # Refused → NA
      
      # 2020 Scale (-9 = Refused, 1 = Male, 2 = Female)
      year == 2020 & gender == 1 ~ 0,  # Male → 0
      year == 2020 & gender == 2 ~ 1,  # Female → 1
      year == 2020 & gender == -9 ~ NA_real_,  # Refused → NA
      
      # 2024 Scale (-1 = Inapplicable, 9 = Refused, 1 = Man, 2 = Woman, 3 = Nonbinary, 4 = Something Else)
      year == 2024 & gender == 1 ~ 0,  # Man → 0
      year == 2024 & gender == 2 ~ 1,  # Woman → 1
      year == 2024 & gender %in% c(3, 4) ~ 2,  # Nonbinary & Something else → 2
      year == 2024 & gender %in% c(-1, 9) ~ NA_real_,  # Inapplicable/Refused → NA
      
      # Catch all unexpected values as NA
      TRUE ~ NA_real_
    ))
}

# Apply function to the combined dataset
all_clean <- recode_gender(all_clean)

# Check distribution of gender after recoding
table(all_clean$gender, useNA = "ifany")
```

Education Level (Ordinal)
Education is typically measured as an ordinal variable, so we align different years to a standard scale:
```{r}
recode_education <- function(df, year) {
  df %>%
    mutate(educ = case_when(
      # 2024 Mapping
      year == 2024 & educ %in% c(1, 2, 3, 4, 5, 6, 7, 8) ~ 1,  # Less than high school
      year == 2024 & educ == 9 ~ 2,  # Some high school (9-12, no diploma)
      year == 2024 & educ == 10 ~ 3,  # High school graduate
      year == 2024 & educ %in% c(11, 12) ~ 4,  # Some college
      year == 2024 & educ == 13 ~ 5,  # Associate degree
      year == 2024 & educ == 14 ~ 6,  # Bachelor's degree
      year == 2024 & educ == 15 ~ 7,  # Master's degree
      year == 2024 & educ == 16 ~ 8,  # Professional/Doctorate degree

      # 2020 Mapping
      year == 2020 & educ == 1 ~ 1,
      year == 2020 & educ == 2 ~ 2,
      year == 2020 & educ == 3 ~ 3,
      year == 2020 & educ %in% c(4, 5) ~ 4,
      year == 2020 & educ == 6 ~ 5,
      year == 2020 & educ == 7 ~ 6,
      year == 2020 & educ == 8 ~ 7,

      # 2016 Mapping
      year == 2016 & educ %in% c(1, 2, 3, 4, 5, 6, 7, 8) ~ 1,  # Less than high school
      year == 2016 & educ == 9 ~ 2,  # Some high school
      year == 2016 & educ == 10 ~ 3,  # High school graduate
      year == 2016 & educ %in% c(11, 12) ~ 4,  # Some college
      year == 2016 & educ == 13 ~ 5,  # Associate degree
      year == 2016 & educ == 14 ~ 6,  # Bachelor's degree
      year == 2016 & educ == 15 ~ 7,  # Master's degree
      year == 2016 & educ == 16 ~ 8,  # Professional/Doctorate degree

      # 2012 Mapping
      year == 2012 & educ %in% c(6, 7, 8) ~ 1,  # Less than high school
      year == 2012 & educ == 9 ~ 2,  # Some high school
      year == 2012 & educ == 10 ~ 3,  # High school graduate
      year == 2012 & educ %in% c(11, 12) ~ 4,  # Some college
      year == 2012 & educ == 13 ~ 5,  # Associate degree
      year == 2012 & educ == 14 ~ 6,  # Bachelor's degree
      year == 2012 & educ == 15 ~ 7,  # Master's degree

      # Handle missing values
      educ %in% c(-9, -8, -1, 95) ~ NA_real_,
      TRUE ~ educ  # Keep other values unchanged
    ))
}

all_clean <- recode_education(all_clean)


```

```{r}
table(all_clean$educ)
```