---
title: "Latino_Voter_Analysis"
output: html_document
date: "2025-03-17"
---

```{r}
# Load necessary libraries
library(haven)  # If dataset is in .sav format
library(carData)    # For VIF (multicollinearity)
#library(DescTools) # For logistic regression diagnostics
library(dplyr)
library(tidyr)
library(ggplot2)

# Load data
df <- read.csv("v2.Latino_Voters_ANES.csv") 


```


Descriptive Statistics
Summarizing key variables: party identification, Republican vote share, religiosity, and political assimilation.

summarize the count and proportion of each category.


```{r}
#change to factor
df$party_id <- factor(df$party_id, 
                      levels = c(1, 2, 3, 4), 
                      labels = c("Democrat", "Republican", "Independent", "Other"))
```

```{r}
df$class <-factor(df$class,
                  levels = c(1,2,3,4),
                  label = c("Lower Class", "Working Class","Middle Class","Upper Class"))
```

```{r}
df$education <-factor(df$education,
                  levels = c(1,2,3,4,5,6,7,8),
                  label = c("Less than HS", "Some HS","HS graduate","Some College","Associate", "Bachelor's", "Master's", "Professional/Doctorate"))
```

```{r}
df$gender <- factor(df$gender,
                    level = c(0,1,2),
                    labels = c("Male", "Female", "Nonbinary/Other"))
```

2024's is the source of the high missing values of republican_vote. This is impacting the dependent variable. 
 I must exclude 2024 from models that use republican_vote. I will filter anes_combined for those analysis. 


```{r}
df_filtered_24 <- df %>%
  filter(!(year == 2024 & is.na(republican_vote)))
```
 
```{r}
# Re-run logistic regression
logit_model <- glm(republican_vote ~ party_id + religiosity + 
                      english_proficiency + political_discussion + 
                      voter_participation + generational_status +
                      age + gender + education + class, 
                   data = df_filtered_24, family = binomial)

summary(logit_model)
```
```{r}
library(ggplot2)
library(broom)
library(forcats)
# Convert model summary to a tidy dataframe
model_summary <- broom::tidy(logit_model, conf.int = TRUE)

# Remove the intercept for better visualization
model_summary <- model_summary %>% filter(term != "(Intercept)")

# Rename variable labels for better readability
model_summary <- model_summary %>%
  mutate(term = fct_recode(term, 
    "Republican" = "party_idRepublican",
    "Independent" = "party_idIndependent",
    "Other Party" = "party_idOther",
    "Political Discussion" = "political_discussion",
    "Bachelor's Degree" = "educationBachelor's",
    "Religiosity" = "religiosity",
    "Female" = "genderFemale",
    "Age" = "age",
    "Voter Participation" = "voter_participation",
    "Associate Degree" = "educationAssociate",
    "Working Class" = "classWorking Class",
    "Master's Degree" = "educationMaster's",
    "English Proficiency" = "english_proficiency",
    "Some College" = "educationSome College",
    "Generational Status" = "generational_status",
    "High School Graduate" = "educationHS graduate",
    "Some High School" = "educationSome HS"
  ))


# Plot coefficients with cleaned labels
ggplot(model_summary, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(color = "blue", size = 3) +  # Plot coefficients
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "black") +  # Add confidence intervals
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Reference line at zero
  coord_flip() +  # Flip coordinates for readability
  theme_minimal() +
  labs(title = "Latino Republican Vote Choice",
       x = "Predictor Variable",
       y = "Coefficient Estimate (Log Odds)")
```
```{r}
library(ggplot2)
library(broom)
library(forcats)
# Convert model summary to a tidy dataframe
model_summary <- broom::tidy(logit_model, conf.int = TRUE)

# Remove the intercept for better visualization
model_summary <- model_summary %>% filter(term != "(Intercept)")

# Rename variable labels for better readability
model_summary <- model_summary %>%
  mutate(term = fct_recode(term, 
    "Republican" = "party_idRepublican",
    "Independent" = "party_idIndependent",
    "Other Party" = "party_idOther",
    "Political Discussion" = "political_discussion",
    "Bachelor's Degree" = "educationBachelor's",
    "Religiosity" = "religiosity",
    "Female" = "genderFemale",
    "Age" = "age",
    "Voter Participation" = "voter_participation",
    "Associate Degree" = "educationAssociate",
    "Working Class" = "classWorking Class",
    "Master's Degree" = "educationMaster's",
    "English Proficiency" = "english_proficiency",
    "Some College" = "educationSome College",
    "Generational Status" = "generational_status",
    "High School Graduate" = "educationHS graduate",
    "Some High School" = "educationSome HS"
  ))


# Plot coefficients with cleaned labels
ggplot(model_summary, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(color = "blue", size = 3) +  # Plot coefficients
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "black") +  # Add confidence intervals
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Reference line at zero
  coord_flip() +  # Flip coordinates for readability
  theme_minimal() +
  labs(title = "Latino Republican Vote Choice",
       x = "Predictor Variable",
       y = "Coefficient Estimate (Log Odds)",
       caption = "Logistic regression estimates of factors influencing Latino Republican vote choice. Coefficients represent log-odds, with positive values indicating an increased likelihood of voting Republican. Black lines show 95% confidence intervals. Data source: ANES 2016-2024."
  ) +
  theme(plot.caption = element_text(size = 10, hjust = 0))
```

# Count of Partisan ID by Year
```{r}
df %>%
  group_by(year, party_id) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = as.factor(year), y = count, fill = party_id)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Partisan Identification Trends (2004-2024)",
       x = "Year", y = "Proportion", fill = "Party ID") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "purple", "gray"))
```
```{r}
df_clean <- df %>% filter(!is.na(party_id))
df_clean %>%
  group_by(year, party_id) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = as.factor(year), y = count, fill = party_id)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Partisan Identification Trends (2012-2024)",
       x = "Year", y = "Proportion", fill = "Party ID") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "purple", "gray"))
```

`

#Republican Vote Share Trends
```{r}
df_clean <- df_clean %>% filter(!is.na(republican_vote))
df_clean%>%
  group_by(year, republican_vote) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = as.factor(year), y = count, fill = as.factor(republican_vote))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Republican Vote Share Among Latino Voters (2012-2024)",
       x = "Year", y = "Proportion of Votes", fill = "Republican Vote (1 = Yes, 0 = No)") +
  theme_minimal()
```
#Religiosity Trends

```{r}
df_clean <- df_clean %>% filter(!is.na(religiosity))
#change to factor
df_clean$religiosity <- factor(df_clean$religiosity, 
                      levels = c(1, 2, 3, 4,5), 
                      labels = c("Every Week", "Almost Every Week", "Once or Twice a Month","A Few Times A Year", "Never" ))
```

```{r}
df_clean %>%
  group_by(year, religiosity) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = as.factor(year), y = count, fill = as.factor(religiosity))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Religiosity Levels of Latino Voters (2012-2024)",
       x = "Year", y = "Proportion", fill = "Religiosity Level") +
  theme_minimal()
```
4. Political Assimilation Trends
Political assimilation is measured using English proficiency, political discussion frequency, and voter participation history.


```{r}

df_clean<- df_clean %>% filter(!is.na(political_assimilation))
df_clean %>%
  group_by(year) %>%
  summarise(
    avg_english_proficiency = mean(english_proficiency, na.rm = TRUE),
    avg_political_discussion = mean(political_discussion, na.rm = TRUE),
    avg_voter_participation = mean(voter_participation, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -year, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = as.factor(year), y = value, group = variable, color = variable)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Political Assimilation Trends (2012-2024)",
       x = "Year", y = "Average Score", color = "Assimilation Measure") +
  theme_minimal()
```
##Cross-Tabulations
Cross-tabulations (contingency tables) help summarize relationships between two categorical variables.

#Republican Vote Choice by Partisan Identification
```{r}
table(df_clean$party_id, df_clean$republican_vote)
prop.table(table(df_clean$party_id, df_clean$republican_vote), margin = 1)  # Row percentages

```

##Logistic Regression- Republican Vote
The model will predict Republican vote choice (1 = Republican, 0 = Non-Republican) using party ID, religiosity, assimilation factors (English proficiency, political discussion, voter participation), and control variables (age, gender, education, etc.).



df_clean$class <-factor(df_clean$class,
                  levels = c(1,2,3,4),
                  label = c("Lower Class", "Working Class","Middle Class","Upper Class"))



df_clean$education <-factor(df_clean$education,
                  levels = c(1,2,3,4,5,6,7,8),
                  label = c("Less than HS", "Some HS","HS graduate","Some College","Associate", "Bachelor's", "Master's", "Professional/Doctorate"))


```{r}
df_clean$gender <- relevel(as.factor(df_clean$gender), ref = "Male")  # If gender is binary
```

```{r}
# Logistic regression model
logit_model <- glm(republican_vote ~ party_id + religiosity + 
                      english_proficiency + political_discussion + 
                      voter_participation + generational_status +
                      age + gender + education + class, 
                   data = df_clean, family = binomial)

# Model summary
summary(logit_model)
```
```{r}
# Convert categorical variables to factors before running the model
df_clean2 <- df_clean %>%
  mutate(party_id = as.factor(party_id),
         gender = as.factor(gender),
         education = as.factor(education),
         class = as.factor(class))  # Convert 'class' explicitly to a factor
```
```{r}
# Re-run logistic regression
logit_model_2 <- glm(republican_vote ~ party_id + religiosity + 
                      english_proficiency + political_discussion + 
                      voter_participation + generational_status +
                      age + gender + education + class, 
                   data = df_clean, family = binomial)

```

```{r}
# Load Stargazer
library(stargazer)

# Generate table
stargazer(logit_model, type = "html", 
          title = "Logistic Regression Results: Latino Republican Vote Choice", 
          dep.var.labels = c("Republican Vote Choice (1 = Republican, 0 = Other)"),
          covariate.labels = c("Party Identification", "Religiosity",
                               "English Proficiency", "Political Discussion", 
                               "Voter Participation", "Generational Status",
                               "Age", "Gender (Female = 1)", "Education Level", "Class"),
          omit.stat = c("aic", "bic"),
          ci = TRUE, 
          single.row = TRUE)
```



##Interaction Model: Partisan Identification × Religiosity
This model examines whether party identification moderates the effect of religiosity on Republican vote choice.

#Party_ID and Religiosity
```{r}
logit_interaction1 <- glm(republican_vote ~ party_id * religiosity + 
                            english_proficiency + political_discussion + 
                            voter_participation + generational_status +
                            age + gender + education + class, 
                          family = binomial, data = df_filtered_24)

summary(logit_interaction1)
```


```{r}
logit_model_poli <- glm(republican_vote ~ party_id * english_proficiency + 
                                   party_id * political_discussion + 
                                   party_id * voter_participation + 
                                   party_id * generational_status +
                                   age + gender + education + class, 
                   data = df_filtered_24, family = binomial)

# Display summary of results
summary(logit_model_poli)
```

```{r}
library(ggeffects)
df_clean<- df %>% filter(!is.na(party_id))

plot1 <- ggpredict(logit_interaction1, terms = c("religiosity", "party_id"))
plot(plot1)
```
#Party Identification x Assimilation
```{r}
logit_interaction2 <- glm(republican_vote ~ party_id * (english_proficiency + 
                            political_discussion + voter_participation) + 
                            religiosity + generational_status +
                            age + gender + education + class, 
                          family = binomial, data = df_filtered_24)

summary(logit_interaction2)
```
Interaction Model: Party Identification × Generational Status
This model tests how the relationship between party identification and Republican vote choice differs across generations.
```{r}
logit_interaction3 <- glm(republican_vote ~ party_id * generational_status + 
                            religiosity + english_proficiency + 
                            political_discussion + voter_participation +
                            age + gender + education + class, 
                          family = binomial, data = df_clean)

summary(logit_interaction3)
```
```{r}
#change to factor

df_filtered_24$generational_status <- factor(df_filtered_24$generational_status, 
                      levels = c(1, 2, 3), 
                      labels = c("Both Parents", "1 Parent Born", "None Born in U.S"))
```
Since the factor levels are strings, filter the dataset using their exact labels:
```{r}
# Subset data by generational status
first_gen <- df_filtered_24 %>% filter(generational_status == "None Born in U.S")  # First generation
second_gen <- df_filtered_24 %>% filter(generational_status == "1 Parent Born")  # Second generation
third_gen <- df_filtered_24 %>% filter(generational_status == "Both Parents")  # Third generation

# Check sample sizes
nrow(first_gen)  # Number of first-gen respondents
nrow(second_gen)  # Number of second-gen respondents
nrow(third_gen)  # Number of third-gen respondents
```
```{r}
# Summarize political assimilation measures by generation
gen_summary <- df_filtered_24 %>%
  group_by(generational_status) %>%
  summarise(
    avg_english = mean(english_proficiency, na.rm = TRUE),
    avg_discussion = mean(political_discussion, na.rm = TRUE),
    avg_participation = mean(voter_participation, na.rm = TRUE)
  )

# Print summary table
print(gen_summary)
```
Later generations should show higher values for political discussion and participation, reflecting greater political assimilation.
First-generation immigrants may have lower English proficiency, reinforcing barriers to political engagement.

```{r}
library(ggplot2)

df_clean %>%
  group_by(generational_status) %>%
  summarise(
    avg_english = mean(english_proficiency, na.rm = TRUE),
    avg_discussion = mean(political_discussion, na.rm = TRUE),
    avg_participation = mean(voter_participation, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(avg_english, avg_discussion, avg_participation),
               names_to = "Assimilation_Measure", values_to = "Average_Score") %>%
  ggplot(aes(x = generational_status, y = Average_Score, fill = Assimilation_Measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Political Assimilation by Generational Status",
       x = "Generational Status", y = "Average Score", fill = "Assimilation Measure") +
  theme_minimal()
```
#Interpretation of Logistic Regression Results Across Generations
This analysis examines how Republican vote choice is influenced by party identification, religiosity, political assimilation (English proficiency, political discussion, voter participation), and demographic controls (age, gender, education, class) among first-, second-, and third-generation Latino voters.

```{r}
# Logistic regression for first-generation Latinos
logit_first_gen <- glm(republican_vote ~ party_id + religiosity + 
                         english_proficiency + political_discussion + 
                         voter_participation + age + gender + education + class, 
                       data = first_gen, family = binomial)

summary(logit_first_gen)

# Logistic regression for second-generation Latinos
logit_second_gen <- glm(republican_vote ~ party_id + religiosity + 
                          english_proficiency + political_discussion + 
                          voter_participation + age + gender + education + class, 
                        data = second_gen, family = binomial)

summary(logit_second_gen)

# Logistic regression for third-generation Latinos
logit_third_gen <- glm(republican_vote ~ party_id + religiosity + 
                         english_proficiency + political_discussion + 
                         voter_participation + age + gender + education + class, 
                       data = third_gen, family = binomial)

summary(logit_third_gen)
```



```{r}
library(dplyr)
library(stargazer) # For formatted regression tables
#library(gt) # Alternative for a cleaner table format
```











``{r}
library(ggeffects)

plot_gen <- ggpredict(logit_interaction_gen, terms = c("party_id", "generational_status"))
plot(plot_gen)
```

```{r}
# Load necessary library
library(ggplot2)

# Create cross-tabulation
table_data <- df %>%
  group_by(party_id, republican_vote) %>%
  summarise(count = n(), .groups = "drop") 

# Heatmap
ggplot(table_data, aes(x = as.factor(party_id), y = as.factor(republican_vote), fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Heatmap of Republican Vote Choice by Partisan ID",
       x = "Party ID", y = "Vote Choice", fill = "Count") +
  theme_minimal()
```

```{r}
# Frequency distributions of key categorical variables
table(df_clean$party_id)  # Partisan identification
table(df_clean$republican_vote)  # Republican vote choice
table(df_clean$religiosity)  # Religiosity levels
table(df_clean$generational_status)  # Generational status

# Cross-tabulations
xtabs(~ party_id + republican_vote, data = df_clean)
xtabs(~ religiosity + republican_vote, data = df_clean)
```

```{r}

library(ggplot2)

# Republican vote share over time
ggplot(df_clean, aes(x=year, fill=as.factor(republican_vote))) +
  geom_bar(position="fill") +
  labs(title="Latino Republican Vote Share Over Time",
       y="Proportion", x="Year") +
  scale_fill_manual(values=c("blue", "red"), labels=c("Non-Republican", "Republican"))

```

Logistic Regression Model
```{r}
# Convert variables to factors
df_clean$republican_vote <- as.factor(df_clean$republican_vote)
df_clean$party_id <- as.factor(df_clean$party_id)
df_clean$religiosity <- as.factor(df_clean$religiosity)
df_clean$generational_status <- as.factor(df_clean$generational_status)

# Logistic Regression Model
logit_model <- glm(republican_vote ~ party_id + religiosity + 
                      english_proficiency + political_discussion + 
                      voter_participation + generational_status + age + gender + 
                      education + class, 
                   data = df_clean, family = binomial)

# Summary of the model
summary(logit_model)
```

```{r}
# Interaction: Party ID x Religiosity
logit_interaction1 <- glm(republican_vote ~ party_id * religiosity + 
                              english_proficiency + political_discussion + 
                              voter_participation + generational_status + age + gender + 
                              education + class, 
                           data = df_clean, family = binomial)

# Interaction: Political Assimilation x Party ID
logit_interaction2 <- glm(republican_vote ~ party_id * (english_proficiency + 
                          political_discussion + voter_participation) + 
                          generational_status + age + gender + education + class, 
                          data = df_clean, family = binomial)

summary(logit_interaction1)
summary(logit_interaction2)
```
Comparative Analysis by Generational Status


```{r}
# Check unique values and counts
table(df_clean$generational_status)

# Visualize distribution
ggplot(df_clean, aes(x = as.factor(generational_status))) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Generational Status", x = "Generational Status", y = "Count") +
  scale_x_discrete(labels = c("1" = "Both parents US", "2" = "One parent US", "3" = "Both parents Foreign"))

```
```{r}
df_clean$generational_status <- as.numeric(as.character(df_clean$generational_status))
str(df_clean$generational_status)
table(df_clean$generational_status, useNA = "always")  # Check if there are NA values
```
```{r}
df_clean$generational_status <- factor(df_clean$generational_status, 
                                 levels = c(1, 2, 3), 
                                 labels = c("Both parents US", "One parent US", "Both parents Foreign"))
```


compare generational status 
Separate Logistic Regression Models
estimates the effect of key independent variables separately for each generational group.

```{r}
# Subset Data by Generational Status
first_gen <- df_clean[df_clean$generational_status == "Both parents Foreign", ]
second_gen <- df_clean[df_clean$generational_status == "One parent US", ]
third_gen <- df_clean[df_clean$generational_status == "Both parents US", ]

# Logistic Regression for First Generation
logit_first_gen <- glm(republican_vote ~ party_id + religiosity + english_proficiency + 
                         political_discussion + voter_participation + age + gender + 
                         education + class, 
                       data = first_gen, family = binomial)

# Logistic Regression for Second Generation
logit_second_gen <- glm(republican_vote ~ party_id + religiosity + english_proficiency + 
                          political_discussion + voter_participation + age + gender + 
                          education + class,
                        data = second_gen, family = binomial)

# Logistic Regression for Third Generation
logit_third_gen <- glm(republican_vote ~ party_id + religiosity + english_proficiency + 
                         political_discussion + voter_participation + age + gender + 
                         education + class, 
                       data = third_gen, family = binomial)

# Compare Model Summaries
summary(logit_first_gen)
summary(logit_second_gen)
summary(logit_third_gen)
```

Look at the coefficients (Estimate) for each variable.
Compare statistical significance (p-values).
Identify differences in predictors across generations.

Interaction Model- Generation Status
test generational differences within a single model by adding an interaction effect.

Why Use Interaction?
Tests whether party ID influences Republican vote choice differently across generations.
More efficient than separate models (avoids loss of statistical power).
If the interaction term (party_id * generational_status) is statistically significant, it suggests that the effect of party ID varies by generation.
```{r}
logit_interaction <- glm(republican_vote ~ party_id * generational_status + religiosity + 
                         english_proficiency + political_discussion + voter_participation + 
                         age + gender + education + class, 
                         data = df_clean, family = binomial)

summary(logit_interaction)
```
none were significant

##Marginal Effects (Probability Differences)
To visualize the impact of generational status on vote probability, use margins.

```{r}

```


```{r}
logit_model <- glm(republican_vote ~ party_id + generational_status + religiosity + 
                   english_proficiency + political_discussion + voter_participation + 
                   age + gender + education + class, 
                   data = df_clean, family = binomial)

summary(logit_model)
```
